\documentclass[]{article}
\usepackage{multicol}
\usepackage{setspace}
\usepackage{mathrsfs}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{hyperref}
\usepackage[labelfont=bf]{caption}
\usepackage{amsmath, amssymb}

%opening
\title{The Reparamaterisation Trick}
\author{Greg Feldmann}

\begin{document}

\maketitle

\begin{abstract}
The reparamatisation trick is used to enable variational autoencoders (VAEs) to be trained with gradient descent.
\end{abstract}

\section{Quick Overview of Variational Autoencoders}
Let $\mathbf{z}$ be a vector of latent variables, $\mathbf{x}$ be a row vector from a dataset $X$, $q_{\phi}(\mathbf{z}|\mathbf{x})$ be the encoder and $p_{\theta}(\mathbf{x}|\mathbf{z})$ be the decoder. The loss function used is referred to as the evidence lower bound (ELBO). ELBO is defined as follows
\begin{equation} \mathcal{L}_{\theta, \phi}(\mathbf{x}) = log(p_{\theta}(\mathbf{x})) - D_{KL}(q_{\phi}(\mathbf{z}|\mathbf{x})||p_{\theta}(\mathbf{z}|\mathbf{x}))\end{equation}
where $D_{KL}$ is the Kullback-Leibler divergence.

\subsection{More on ELBO}

\url{https://www.zinkov.com/posts/2018-11-02-decomposing-the-elbo/}

\section{Optimising ELBO with SGD}
To optimise our VAE, we need to be able to take gradients of the expected value of ELBO with respect to the network weights $\theta$ and $\phi$. This is easy enough for $\theta$:
\begin{equation} \nabla_{\theta}\mathbb{E}_{q_{\phi}(\mathbf{z}|\mathbf{x})}[\mathcal{L}_{\theta, \phi}(\mathbf{x})] = \mathbb{E}_{q_{\phi}(\mathbf{z}|\mathbf{x})}[\nabla_{\theta} \mathcal{L}_{\theta, \phi}(\mathbf{x})]
\end{equation}
Once the grad operator is inside the expectation, all we have to do is approximate the expectation with a sample. 
\newline
\newline
Things are trickier with $\phi$. As the expectation assumes that $p(\mathbf{z}|\mathbf{x}) = q_{\phi}(\mathbf{z}|\mathbf{x})$, we cannot simply move $\nabla_{\phi}$ in and out of the expectation arbitrarily.
\begin{equation} \nabla_{\phi}\mathbb{E}_{q_{\phi}(\mathbf{z}|\mathbf{x})}[\mathcal{L}_{\theta, \phi}(\mathbf{x})] \neq \mathbb{E}_{q_{\phi}(\mathbf{z}|\mathbf{x})}[\nabla_{\phi} \mathcal{L}_{\theta, \phi}(\mathbf{x})]
\end{equation}
That the left and right hand sides are not equal can be seen as follows:
\begin{equation} \nabla_{\phi}\mathbb{E}_{q_{\phi}(\mathbf{z}|\mathbf{x})}[\mathcal{L}_{\theta, \phi}(\mathbf{x})] = \nabla_{\phi} \int q_{\phi}(\mathbf{z}|\mathbf{x}) \mathcal{L}_{\theta, \phi}(\mathbf{x}) d\mathbf{x}
\end{equation}

\begin{equation}
= \int \nabla_{\phi}(q_{\phi}(\mathbf{z}|\mathbf{x}) \mathcal{L}_{\theta, \phi}(\mathbf{x})) d\mathbf{x}
\end{equation}

\begin{equation}
= \int [q_{\phi}(\mathbf{z}|\mathbf{x})(\nabla_{\phi} \mathcal{L}_{\theta, \phi}(\mathbf{x})) + \mathcal{L}_{\theta, \phi}(\mathbf{x}) (\nabla_{\phi} q_{\phi}(\mathbf{z}|\mathbf{x}))] d\mathbf{x}
\end{equation}

\begin{equation}
= \int q_{\phi}(\mathbf{z}|\mathbf{x})(\nabla_{\phi} \mathcal{L}_{\theta, \phi}(\mathbf{x})) d\mathbf{x} + \int \mathcal{L}_{\theta, \phi}(\mathbf{x}) (\nabla_{\phi} q_{\phi}(\mathbf{z}|\mathbf{x})) d\mathbf{x}
\end{equation}

\begin{equation}
= \mathbb{E}_{q_{\phi}(\mathbf{z}|\mathbf{x})}[\nabla_{\phi} \mathcal{L}_{\theta, \phi}(\mathbf{x})] + \int \mathcal{L}_{\theta, \phi}(\mathbf{x}) (\nabla_{\phi} q_{\phi}(\mathbf{z}|\mathbf{x})) d\mathbf{x}
\end{equation}
This isn't in a form that we can easily handle. In particular, $\mathbf{z} \sim q_{\phi}(\mathbf{z}|\mathbf{x})$, meaning $\mathbf{z}$ is stochastic rather than deterministic. This is where the reparamaterisation trick comes into play. We can make $\mathbf{z}$ deterministic by making it a function of a stochastic variable, $\epsilon$:

\begin{equation}
\mathbf{z} = g(\epsilon, \phi, \mathbf{x})
\end{equation}
After replacing $\mathbf{z}$ with $g(\epsilon, \phi, \mathbf{x})$, the expectation is then taken over the distribution of $\epsilon$. The stochasticity is now separated from $\theta$, allowing us to calculate $ \nabla_{\phi}\mathbb{E}_{q_{\phi}(\mathbf{z}|\mathbf{x})}[\mathcal{L}_{\theta, \phi}(\mathbf{x})]
$ as follows:
\begin{equation}
\nabla_{\phi}\mathbb{E}_{q_{\phi}(\mathbf{z}|\mathbf{x})}[\mathcal{L}_{\theta, \phi}(\mathbf{x})] = \nabla_{\phi}\mathbb{E}_{p(\epsilon)}[\mathcal{L}_{\theta, \phi}(\mathbf{x})]
\end{equation}

\begin{equation}
= \mathbb{E}_{p(\epsilon)}[\nabla_{\phi}\mathcal{L}_{\theta, \phi}(\mathbf{x})]
\end{equation}

\nocite{*}
\bibliography{refs}
\bibliographystyle{amsplain}
\end{document}
